\documentclass[10pt]{scrbook}
\usepackage{geometry}
\usepackage{graphicx}
\usepackage{import}
\usepackage{xcolor}
\usepackage{listings}
\usepackage{caption}
\usepackage{hyperref}
\usepackage{tikz}
\usetikzlibrary{calc,positioning,shapes.geometric}

\geometry{a4paper}
\setlength{\parindent}{0pt}

%\DeclareCaptionFont{white}{\color{white}}
%DeclareCaptionFormat{listing}{\colorbox[cmyk]{0.43, 0.35, 0.35,0.01}
\DeclareCaptionFormat{listing}{\colorbox{black!15}
  {\parbox{5cm}{#1#2#3}}}
\captionsetup[lstlisting]{format=listing,
%labelfont=white,textfont=white, 
singlelinecheck=false, margin=0pt, font={bf,footnotesize}}

\begin{document}
\title{OpenPEARL - Platform Manual for Linux Systems}
\author{R. M\"uller}

\lstnewenvironment{CppCode} {
    \lstset{numbers=left,
            title={C++},
            frame=tlrb,
	    breaklines = true,
	    belowcaptionskip=4pt
    }
}%
{}%

\lstnewenvironment{PEARLCode} {
    \lstset{numbers=left,
            title={PEARL},
            frame=tlrb,
	    breaklines = true,
	    belowcaptionskip=4pt
    }
}%
{}%

\pagestyle{myheadings}
\markboth {OpenPEARL Platform Manual Linux}
          {OpenPEARL Platform Manual Linux}

\maketitle

\tableofcontents

%\input{log.tex}
%\input{interrupt.tex}
%\input{inOutput.tex}

\chapter{Installation}
The installation must be done via the git repository of the OpenPEARL project.
There are several software packages required on your linux PC. 
The repository contains an install script (\verb|installPackages|),
which identifies your
linux distribution by content of the file \verb|/etc/os-release|.

For details of the additional packages study the installation script.

\paragraph{Prerequisites}
\begin{itemize}
\item you have access to your linux system as user and root
\item you have an internet connection
\item you have a git client installed
\end{itemize}

\paragraph{Perform the following steps:}
\begin{enumerate}
\item login as normal user
\item create a working directory for the installation. \\
   e.g. /home/userx/OpenPEARL\\
    \verb|mkdir /home/userx/OpenPEARL|
\item set you current working directory to this point \\
    \verb|cd /home/userx/OpenPEARL|
\item obtain a read only copy the repository\\
    \verb|git clone git://git.code.sf.net/p/openpearl/code .|\\
    regard the point at the end of the command
\item set the working directory to the top of the real content
     in your working copy 
     (we are sorry about this unnecessary directory level): \\
    \verb|cd openpearl-code|
\item get administration priviledges\\
    \verb|sudo| or \verb|su|
\item run the script:\\
    \verb|./installPackages|\\
    and wait for completion of the installation.
    This takes about 6 minutes at an RaspberryPi 3B
\item still as administrator, execute\\
    \verb|make defconfig|\\
    \verb|make prepare|\\
    \verb|make install|\\
    this takes about 20 minutes at a Rspberry Pi 3B
\end{enumerate}

\paragraph{Verify Installation:}
Work as normal user.


\begin{enumerate}
\item \verb|cd /home/userx/OpenPEARL/openpearl-code/demos|
\item build some of the provided demonstration programs\\
   \verb|prl <demo>.prl|\\
   where \verb|<demo>| is Hello or sched\_demo
\item run the demonstration programm with \\
   \verb|./Hello| or \verb|./sched_demo|.\\
   The programs should pront some messages and terminate themself.
\end{enumerate}


\chapter{Configuration}
\section{Available Options}
\begin{itemize}
\item Target platform: standard linux PC or Raspberry Pi
\item additional device support like PEAK CAN-adapter, I2C-devices,...
\end{itemize}
For details see the {\em HELP} pages in the \verb|make menuconfig| 
entries.

\section{Rebuild of the OpenPEARL Environment}
The tuning the installation must be done as administrator:
\begin{enumerate}
\item \verb|cd /home/userx/OpenPEARL/openpearl-code|
\item change the option upon your demands:\\
      \verb|make menuconfig| 
\item \verb|make install|
\end{enumerate}

\section{Configuration Option for the individual Application}
\import{../../common/platformManual/}{configurationCommon.tex}
\import{./}{log.tex}
\import{./}{pearlrc.tex}

\chapter{Release Information}
\section{Supported Language Elements}
The OpenPEARL project aims to support all language elements  as
described in the language report.

This version does not support:
\begin{itemize}
\item STRUCT
\item TYPE
\item TFU  
\item ...??? fehlt noch was
\item Semantic analysis for good diagnostic error messages.
     In case of problems error messages from the c++ compiler may appear. 
\item semantic analysis for program quality analysis (e.g. unused variables)
\end{itemize}

\section{Restrictions for privileged and unprivileded users}
\begin{itemize}
\item The priority based scheduler of the linux system is not
   accessible for normal users. Thus application will operate for 
   normal users with
   the usual completely fair scheduler. The priority values are 
   not regarded in the scheduling.
\item If the application is run with root priviledges, the socalled 
  round-robin scheduler is used. This provides only 100 priority values.
  For administrative reasons, some priorities are clipped by the OpenPEARL
  system.
  The remaining priorities are: 1-97 and 255
\end{itemize}

\chapter{Supported Devices}
\import{../../common/platformManual/}{devicesCommon.tex}
\import{./}{devices.tex}

\chapter{Supported Interrupt Sources}
\import{../../common/platformManual/}{interruptCommon.tex}
\import{./}{interrupts.tex}

\chapter{Available PEARL Signals}
The OpenPEARL signals are organized in some groups.

\import{../../common/}{signallist.tex}


\chapter{How-To Add new System Elements}
\import{../../common/platformManual/}{addDeviceCommon.tex}
\import{./}{addDevice.tex}
\import{../../common/platformManual/}{addInterruptCommon.tex}
\import{./}{addInterrupt.tex}
\import{../../common/platformManual/}{addSignalCommon.tex}
\end{document}

