\section{LogFile}
Provides the capability to specify the log file name.
The LogFile needs another system device, which provides the access to a 
disc.

Synopsis: \texttt{LogFile(filename)}

\begin{description}
\item[filename]  is the name of the fil, which collects all log messages
   from the application.
\end{description}

\paragraph{Note:} The configuration element LogFile needs a disc type device
  the for the access to the phyical storage.
 

Sample Usage:
\begin{PEARLCode}
disc: Disc('/tmp/' ,10); ! linux specific disc definition
Log('EW') --- LogFile('pearl.log') --- disc;
\end{PEARLCode}

\section{I2c-Device: LM75}
The sensor delivers the current temperature via an I2C-bus interface.

Synopsis: \texttt{LM75(address)}

\begin{description}
\item[address] is the I2C adress of the sensor on the I2C bus.
   Valid adresses are: 48'B4, '49'B4, '4A'B4, '4B'B4,
            '4C'B4, '4D'B4 , '4E'B4, '4F'B4
\end{description}

\paragraph{Note:}\ 
\begin{itemize}
\item the temperatur is returned in thenth of degree centigrade as a 
    FIXED(15) value
\item The chip needs some time to update the temperature value. A gap of
   $\approx 40\mu s$ should be kept by the application between two read cycles.
\item The LM75 device needs an I2C-bus device to operate
\end{itemize}


Sample Usage:
\begin{PEARLCode}
...
SYSTEM;
   ! linux specific i2c bus interface definition
   ic2bus1: I2CBUS('/dev/i2c-0');

   ! connect the sensor to the I2C bus
   temp: LM75('48'B4) --- i2cbus1;
...
PROBLEM;
   SPC temperatur DATION IN SYSTEM BASIC FIXED(15);
   DCL uTemperatur DATION IN BASIC FIXED(15) CREATED(temperatur);
...
t1: TASK;
   DCL tValue FIXED(15);
...
   OPEN uTemperatur;
   TAKE tValue FROM uTemperatur;
   PUT tvalue/10,'.',tValue REM 10 TO console BY F(3),A,F(1), SKIP;
...
   CLOSE uTemperatur;
END;
...
\end{PEARLCode}

\section{I2c-Device: PCF8574}
This device acts as port expander with 8 bit numbered from 7
for the most significant bit down to 0.
The decive operates via an I2C bus. 
Individual bits may be ether as input or as output.
Input and output are treated in different system elements.

Synopsis: PCF8574In(address, start, width) \\
\ \ \ \ PCF8574Out(address, start, width)

The access of these devices must be done with BIT-types of the same size
as denoted by width.

\begin{description}
\item[address] is the I2C adress of the sensor on the I2C bus.
   Valid adresses are: '20'B4, '21'B4, '22'B4, '23'B4,
            '24'B4, '25'B4 , '26'B4, '27'B4,
            '38'B4, '39'B4, '3A'B4, '3B'B4, '3C'B4, '3D'B4 '3E'B4, '3F'B4
\item[start] is the starting bit number, starting at the most signifigant bit 
    number. 
\item[width] the number of consecutive bits to use
      towards the least significant direction
\end{description}

\paragraph{Notes:}\ 
\begin{itemize}
\item The PCF8574In device is read only
\item The PCF8574Out device is write only
\item Example: PCF8574In(5,4) will use bits 5,4,3 and 2 
\end{itemize}

\begin{PEARLCode}
...
SYSTEM;
   ! linux specific i2c bus interface definition
   ic2bus1: I2CBUS('/dev/i2c-0');

   ! connect the led to bit 7 of the port expander
   ! via the I2C bus
   led: PCF8574Out('21'B4,7,1) --- i2cbus1;

   ! connect a switch to bit 6 of the port expander
   ! via the I2C bus
   switch: PCF8574In('21'B4,6,1) --- i2cbus1;

   ! connect a stepper motor to bit 3,2,1,0 of the port expander
   ! via the I2C bus
   stepper: PCF8574Out('21'B4,7,1) --- i2cbus1;

PROBLEM;
   SPC led DATION OUT SYSTEM BASIC BIT(1);
   DCL uLed DATION OUT BASIC BIT(1) CREATED (led);

   SPC switch DATION IN SYSTEM BASIC BIT(1);
   DCL uSwitch DATION IN BASIC BIT(1) CREATED (switch);

   SPC stepper DATION OUT SYSTEM BASIC BIT(4);
   DCL uStepper DATION OUT BASIC BIT(4) CREATED (stepper);

t1: TASK;
   DCL currentSetting BIT(1);

   OPEN uLed;
   OPEN uSwitch;
   OPEN uStepper;
   SEND '1'B1 TO uLed;  ! switch on
   SEND '0'B1 TO uLed;  ! switch off

   TAKE currentSetting FROM uSwitch;
   SEND currentSetting TO uLed;  ! echo the switch position

   SEND '1010'B1 TO uStepper;
 
   CLOSE uLed;
   CLOSE uSwitch;
   CLOSE uStepper;
...
END;
...
\end{PEARLCode}

\section{I2c-Device: ADS1015SE}
The sensor provides four single ended analog inputs via an I2C bus.
The device  has a programmable gain unit, with modifies the sesibility
of the external voltage by a muliplication by a selecteable factor.

Synopsis: \texttt{ADS1015SE(address, channal, gain)}

The device returns a FIXED(15) value reflecting the external voltage as ratio
of the $\frac{returned value}{32767}*U_{ref}$.
The reference voltage $U_{ref}$ is usually the supply voltage of 3.3V or 5V.

\begin{description}
\item[address] is the I2C adress of the sensor on the I2C bus.
   Valid adresses are: '48'B4, '49'B4, '4A'B4, '4B'B4
\item[channel] is the number of the selected input channel. 
   Valid values are 0,1,2,3
\item[gain] is the selected gain.
   The result of the multiplication must not exceed the supply voltage.

  \begin{tabular}{l|l}
   gain & factor\\
   \hline 
   0 & 2/3 \\
   1 & 1 \\
   2 & 2 \\
   3 & 4 \\
   4 & 8 \\
   8 & 16 \\
   \end{tabular} 

\end{description}

\paragraph{Notes:} \ 
\begin{itemize}
\item The driver supports also the ADS1012 device.
\item The device is read only
\end{itemize}

Sample Usage:
\begin{PEARLCode}
...
SYSTEM;
   ! linux specific i2c bus interface definition
   ic2bus1: I2CBUS('/dev/i2c-0');

   ! connect the sensor to the I2C bus at 
   ! address 48 (hexadecimal)
   ! select channel 0 and internal gain of 1
   voltage1: ADS1015SE('48'B4,0,1) --- i2cbus1;

PROBLEM;
   SPC voltage1 DATION IN SYSTEM BASIC FIXED(15);
   DCL u1 DATION IN BASIC FIXED(15) CREATED(voltage1);

t1: TASK;
   DCL uFixed FIXED(15);
   DCL u FLOAT;
 
   OPEN u1;
   TAKE uFixed FROM u1;
   u := TOFLOAT(uFixed)/32767.0 * 3.3;
   PUT 'voltage:', u, 'V' TO console BY A,X,F(5,3),A,SKIP;
   CLOSE u1;
END;
...
\end{PEARLCode}


