\chapter{Bolt}
PEARL requires a BOLT implementation with simultaneous locking
mechanism.

A single PEARL BOLT is
a state variable and a counter --- 
and for diagnosic purposes a pointer to the bolt's name.

All public bolt operations lock the class mutex to avoid race conditions 
with other tasks bolt operations.

On freeing or leaving a PEARL bolt variable,
 the wait queue content is checked if a 
task may become continued. The wait queue is searched in order of the tasks 
priority. 

To block a task the class must provided the methods {\em block} and 
{\em unblock}.

\section{Declaration}
The declaration of a PEARL bolt variable is done without preset value..

\begin{PEARLCode}
DCL b0 BOLT;
DCL b1 BOLT;
DCL b2 BOLT;
SPC b2 BOLT;
\end{PEARLCode}

The class {\em Bolt} provides two macros for definition and 
specification of a bolt variable.
\begin{description}
\item[DCLBOLT] takes 1 parameter: the bolt's name.
\item[SPCBOLT] takes 1 parameter:  the bolt's name.
\end{description}

\begin{CppCode}
DCLBOLT(b0);
DCLBOLT(b1);
DCLBOLT(b2);
SPCBOLT(b2);
\end{CppCode}

\section{Operations}
The class {\em Bolt} provides the \verb|static| operations 
{\em enter()}, {\em leave()}, {\em reserve()} and  {\em free()}.

PEARL provides the simultaneous locking pattern on bolt variables.
This leads to the interface of the operations:
\begin{CppCode}
void Bolt::enter(Task *me, int nbrOfBolts, Bolt **bolts);
void Bolt::leave(Task *me, int nbrOfBolts, Bolt **bolts);
void Bolt::reserve(Task *me, int nbrOfBolts, Bolt **bolts);
void Bolt::free(Task *me, int nbrOfBolts, Bolt **bolts);
\end{CppCode}

\section{Sample Usage}

\begin{PEARLCode}
ENTER b0,b1,b2;
\end{PEARLCode}

\begin{CppCode}
{
   static Bolt * bolts[] = {&b0,&b1,&b2};
   Bolt::enter(me,
               sizeof(bolts[])/sizeof(bolts[0]),
               &bolts);
}
\end{CppCode}

The new block allows the reuse of the identifier {\em bolts} in multiple
{\em enter}, {\em leave}, {\em reserve} and {\em free} operations
 in the same module.
The static storage declaration provides the initialization before run time.
The size calculation should be executed by the C++-compiler.

\section{Error Handling}
If a LEAVE or FREE is executed upon an non ENTERed or RESERVEd
bolt variable the {\em BoltStateSignal} is induced.
